---
title: "Introduction to Maps in R"
author: "Jill Thomley"
date: '`r format(Sys.time(), "%A, %B %d, %Y @ %I:%M %p")`'
output: ioslides_presentation
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = NA
)
```

## Packages

```{r}
library(tidyverse)
```

We need `tidyverse` for data manipulation and plotting.

```{r}
library(maps)
```

Package containing geospatial data and functions for R.

```{r}
library(mapdata)
```

The mapdata package extends maps with more datasets.

<hr>

Let's go into the **maps** package and see what data is available!


# Some Mapping Basics

## Dataset `usa`

```{r}
usa <- map_data("usa")
head(usa, 2)
```

```{r}
str(usa)
```


## Regions

```{r}
usa %>% count(region)
```


## Mapping `usa`

Plot coordinates as paths in the x-y plane using [`geom_polygon`](https://ggplot2.tidyverse.org/reference/geom_polygon.html). Group divides the points into non-contiguous parts, like islands.

```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
ggplot(usa) +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "black", fill = "skyblue")
```


## A Different Projection

Using the [Mercator](https://desktop.arcgis.com/en/arcmap/latest/map/projections/mercator.htm) projection method.

```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
ggplot(usa) +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "black", fill = "skyblue") +
  coord_map()
```


## Yet Another Projection

Using the [Mollweide](https://desktop.arcgis.com/en/arcmap/latest/map/projections/mollweide.htm) projection method.

```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
ggplot(usa) +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "black", fill = "skyblue") +
  coord_map(projection = "mollweide")
```


## Dataset `state`

```{r}
state <- map_data("state")
head(state, 12)
```


## Mapping `state`

```{r, fig.height = 3.5, fig.align = "center"}
ggplot(state) +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "black", fill = "lightgreen") +
  coord_map()
```


## Dataset `county`

```{r}
county <- map_data("county")
head(county, 10)
```


## Mapping `county`

```{r, fig.height = 3.5, fig.align = "center"}
ggplot(county) +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "black", fill = "lightgreen") +
  coord_map()
```


## Filtering `state` for NC

```{r, fig.height = 3, fig.align = "center"}
filter(state, region == "north carolina") %>% 
  ggplot() +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "black", fill = "coral") +
  coord_map()
```


## Filtering `county` for NC

```{r, fig.height = 3, fig.align = "center"}
filter(county, region == "north carolina") %>% 
  ggplot() +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "black", fill = "coral") +
  coord_map()
```


## Filtering `world` for USA

```{r}
world <- map_data("world")
str(world)
world_usa <- filter(map_data("world"), region == "USA")
```


## Mapping `world_usa`

Oops, there's a bit of Alaska on the wrong side...

```{r, fig.height = 3, fig.align = "center"}
ggplot(world_usa) +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "black", fill = "plum") +
  coord_map() +
  scale_x_continuous(breaks = seq(-160, 180, by = 20))
```


## Adjusting Our X-Axis

Change orientation from default to center at longitude 180.

```{r, fig.height = 3, fig.align = "center"}
ggplot(world_usa) +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "black", fill = "plum") +
  coord_map(orientation = c(90, 180, 0)) +
  scale_x_continuous(breaks = seq(-160, 180, by = 20))
```


## A Different Adjustment 

```{r, fig.height = 3, fig.align = "center"}
world_usa %>%
  mutate(long_adj = ifelse(long > 0, long - 360, long)) %>% 
  ggplot() +
  geom_polygon(aes(x = long_adj, y = lat, group = group),
               color = "black", fill = "lightgreen") +
  coord_map() +
  scale_x_continuous(breaks = seq(-200, 0, by = 20))
```


## The `mollweide` Projection

```{r, fig.height = 3, fig.align = "center"}
worldmap <- world_usa %>%
  mutate(long_adj = ifelse(long > 0, long - 360, long)) %>% 
  ggplot() +
  geom_polygon(aes(x = long_adj, y = lat, group = group),
               color = "black", fill = "lightgreen")
worldmap + coord_map(projection = "mollweide")
```


## The `ortho` Projection

We can store the basic map as an object and then add different customizations without repeating the basic mapping code. This is true for any ggplot2 plot.

```{r, fig.height = 3, fig.align = "center"}
worldmap + coord_map(projection = "ortho")
```


## Regions in `world_usa`

Notice we have both regions and subregions in this set.

```{r}
str(world_usa)
```

## Region and Subregions

The count `n` is the number of coordinates for the subregion.

```{r}
tail(world_usa %>% count(region, subregion), 10)
```


## Mapping Hawaii

In this dataset we filter for a US state using `subregion`.

```{r, fig.height = 2.5, fig.align = "center"}
filter(world_usa, subregion == "Hawaii") %>% 
  ggplot() +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "black", fill = "turquoise") +
  coord_map()
```


# Removing the Axes

## Ditch the Axes!

We usually do not need the latitude and longitude axes on our maps, so we can use theme elements to get rid of them.

```{r}
# Theme Credit: http://eriqande.github.io/rep-res-web/lectures/
# making-maps-with-R.html (w/ the addition of panel.background)
# See also: https://ggplot2.tidyverse.org/reference/theme.html

ditch_the_axes <- theme(
  axis.title = element_blank(),
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  panel.background = element_blank()
)
```


## Voila, a plot with no axes!

```{r, fig.height = 3, fig.align = "center"}
ncmap <- filter(county, region == "north carolina") %>% 
  ggplot() +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "black", fill = "lightgoldenrod") +
  coord_map()
ncmap + ditch_the_axes
```


## A More Elaborate Version

How is this similar to and different from the other version?

```{r}
no_axes_alt <- theme(
  panel.grid.major = element_line(colour = "transparent"),
  panel.grid.minor = element_blank(),
  axis.title.x = element_blank(),
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.title.y = element_blank(),
  axis.text.y = element_blank(),
  axis.ticks.y = element_blank(),
  panel.background = element_blank(),
  panel.border = element_blank(),
  plot.background = element_blank()
  )
```


## Another plot with no axes!

Again we use our stored map object and add customization.

```{r, fig.height = 3, fig.align = "center"}
ncmap + no_axes_alt
```


# Plotting Data on a Map

## Boone on the NC Map

```{r, fig.height = 2.5}
ncmap +
  geom_point(aes(x = -81.6746, y = 36.2168), 
             size = 1.25, shape = 8, color = "red") +
  labs(title = "Location of Boone, North Carolina",
       caption = "36.22째 N, 81.67째 W") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ditch_the_axes
```


## Boone on the NC Map

```{r, fig.height = 2.5}
ncmap +
  geom_point(aes(x = -81.6746, y = 36.2168), 
             size = 1.25, shape = 8, color = "red") +
  labs(title = "Location of Boone, North Carolina",
       caption = "36.22째 N, 81.67째 W") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ditch_the_axes
```


## Places Dr. Thomley Has Lived

```{r, livemap, eval = FALSE}
places = tibble(lats = c(41.6528, 42.3736, 42.6526, 36.2168),
                longs = c(-83.5379, -71.1097, -73.7562, -81.6746),
                names = c("Toledo", "Cambridge", "Albany", "Boone"))

ggplot(state) +
  geom_polygon(aes(x = long, y = lat, group = group),
               color = "white", fill = "thistle") +
  geom_point(data = places, aes(x = longs, y = lats),
             size = 2, shape = 8, color = "purple") +
  geom_label(data = places, aes(x = longs, y = lats, label = names), 
             hjust = c("right", "right", "left", "right"),
             nudge_x = c(-1, -0.5, 0.5, -1),             
             vjust = c("middle", "top", "bottom", "middle"),
             nudge_y = c(0, -0.5, 0.5, 0)) +
  coord_map() +
  ditch_the_axes
```

## The Map

```{r, ref.label = I("livemap"), echo = FALSE, eval = TRUE}

```


# Resources

##

* http://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html
* https://github.com/UrbanInstitute/urbnmapr
* https://socviz.co/maps.html

